---
title: "Ciccolini_Lab1"
author: "Skylar Ciccolini"
date: "2025-08-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Read in Data {.tabset}

### 1.2.3 Reading in the gapminder data, reversing country order, and displaying first 15 countries.

```{r}

library(gapminder)

gapminder_sorted <- sort(gapminder$country, decreasing = TRUE)

head(gapminder_sorted, n = 15)

str(gapminder)

```

This might not work as well if the country column is a character instead of a factor.

### 1.2.4 Calculating summary statistics for Life expectancy, Population, and GDPperCapita.

```{r}
# first, mean
mean(gapminder$lifeExp)

mean(gapminder$pop)

mean(gapminder$gdpPercap)

# second, median
median(gapminder$lifeExp)

median(gapminder$pop)

median(gapminder$gdpPercap)

# third, stand dev
sd(gapminder$lifeExp)

sd(gapminder$pop)

sd(gapminder$gdpPercap)


```

### 1.2.5 Calculate summary statistics in a single dataframe for Life expectancy, Population, and GDPperCapita for each continent.

```{r}
LE_country <- aggregate(gapminder[ , "lifeExp"], by = gapminder["country"],
                        FUN = mean, na.action = rm)

pop_country <- aggregate(gapminder[ , "pop"], by = gapminder["country"],
                        FUN = mean, na.action = rm)

gdp_country <- aggregate(gapminder[ , "gdpPercap"], by = gapminder["country"],
                        FUN = mean, na.action = rm)


# now put them together
gapminder_summary <- cbind(LE_country, pop_country$pop, gdp_country$gdpPercap)

head(gapminder_summary)

```

### 1.2.6 Use the first 5 records of the global average Life Expectancy over time.

```{r}
avg_LE <- aggregate(gapminder[ , "lifeExp"], by = gapminder["year"],
                    FUN = mean, na.action = rm)

# and now to plot
plot(avg_LE[1:5, ])

# make it cooler
plot(avg_LE[1:5, ], col = "orange", bg = "pink", pch = 10, cex = 3)

```

### 1.2.7 Check out a new dataset from datasets package

```{r}
str(chickwts)

dim(chickwts)

head(chickwts)

```


### 1.2.8 Calculate some summary stats for faithful dataset

```{r}
# first, mean
mean(chickwts$weight)

# second, median
median(chickwts$weight)

# third, standard dev
sd(chickwts$weight)

```


### 1.2.9 Calculate summary statistics for chick weight by feed type

```{r}
chickwts_mean <- aggregate(chickwts[ , "weight"], by = chickwts["feed"],
                           FUN = mean)

head(chickwts_mean)


```


### 1.2.10 Create two plots (one with grouped data, one without)

```{r}
# with grouped data
plot(chickwts_mean, main = "Average Chick Weights by Feed Type",
       xlab = "Feed Type", ylab = "Weight", col = "red", cex = 2)

# without grouped data
plot(chickwts, main = "Chick Weights by Feed Type",
     xlab = "Weight", ylab = "Feed Type")

```


### 1.2.11 Use the ls() to see objects in workspace, remove objects

```{r}
ls()

rm(list = ls())

ls()

```





























